Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-08 text/ms.tex
Edits to text

Black swans are statistically improbable events that nonetheless occur --- often with profound consequences. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). Black swans are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors. Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such. The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies --- particularly as the frequency and magnitude of climatic extremes increase over the next century.

@@

Black swans are unexpected extreme events with potentially dramatic consequences \citep{taleb2007,sornette2009}. One of the most striking black swans in ecology is the asteroid collision marking the end-Cretaceous mass extinction 65 million years ago \citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with shifts in mean temperature, are expected to cause the greatest ecological and societal damage \citep{ipcc2012}. But, while extremes in the physical environment such as wave height, storm severity, and temperature are frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent to which ecological systems buffer or suffer from black swans \citep{nunez2012}.

@@

There are two key reasons why we may find little evidence of ecological black swans. First, they might not exist in higher taxa. Indeed, the majority of model fitting and risk forecasting assumes that population dynamics are not heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively, heavy-tailed dynamics might exist, but our ability to detect them requires further development of statistical tools. One such tool is the generalized extreme value distribution, which has been applied to environmental data \citep[e.g.][]{katz2005}. This distribution describes the most extreme event per time interval (e.g.~heaviest rainfall per year), but requires time series that are sufficiently long to be condensed into time intervals. Another statistical tool involves fitting a catastrophic mixture distribution in a state space model to quantify the probability that population events are extreme, although this is also data intensive \citep{ward2007} A third tool is to compare the fits of thin- and heavy-tailed distributions with the Akaike Information Critera (AIC) TODO CHECK \citep{halley2002}, but this analysis did not quantify the probability of black swans or allow for population dynamics.

Here we develop a framework for identifying ecological black swans in the process noise of population dynamics, i.e.\ the stochastic jumps in abundance from one time step to the next. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series. We apply our model to \NPops\ of populations from four taxonomic classes. We estimate the frequency and strength of black swans in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
%We find a high probability of ecological
%black swans across a wide range of taxa but in a limited number of populations.
%When they occur, extreme climate, predation, parasites, and their interactions
%are common documented causes. We find little evidence that intrinsic
%life-history characteristics are related to the probability of black swans.

@@

To obtain estimates of the probability and magnitude of black swans, we fit population dynamic models to abundance time series from around the world. For each population, we estimated the shape of the process noise tails by measuring the degrees of freedom ($\nu$) of a Student t-distribution.

@@

We selected abundance time series from the Global Population Dynamics Database (GPDD) \citep{gpdd2010}, which contains nearly 5000 time series of abundance from $\sim$1000 species and $\sim$100 taxonomic orders. We filtered the data (Supporting Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values \citep[\textit{sensu}][]{brook2006a}. TODO CONTEXTUALIZE THE QUANTITY OF INTERPOLATION Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps~time steps (range of \minTimeSteps--\maxTimeSteps) (Table S1, Fig.~\ref{fig:all-ts}).

@@

By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b). For example, at $\nu = 2$, the probability of drawing a value more than five standard deviations below the mean is $0.018$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$) TODO CHECK. As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).

@@

We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}. This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity. In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to weaker and stronger priors (Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).

@@

% "Perhaps then you could say "we examined all time series with published explanations of why the black swan events occurred"
TODO STATE HOW THESE 7 WERE CHOSEN
TODO QUANTIFY UPS AND DOWNS --- SKEW
We assembled the documented causes of black-swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table 1). The majority of documented events were downward black swans and involved a series of unfortunate events. For example, a shortage of nest sites for the shag long-necked bird (TODO SCI.\ NAME) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black swan for a water vole population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.

@@

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~1). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

@@

The following supporting information is available online for this article:\\
Figures S1--S11\\
Table S1\\
Example Stan code for a heavy-tailed Gompertz model and multilevel beta
regression\\
The GPDD IDs used in our analysis\\
All code and data to recreate our analysis are available at:\\
\url{https://github.com/seananderson/heavy-tails}

>>>>

Black swans are statistically improbable events that nonetheless occur --- often with profound consequences. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare occurring in \overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). Black swans were not explained by any life-history covariates, but tended to be driven by external perturbations such as climate, severe winters, predator or parasite cycles, and the combined effects of multiple factors. Black swans were more frequently detected for populations with longer time series and lower levels of process noise; for shorter and noisier time series, our simulations suggested black-swan dynamics are often not identifiable as such. The presence of black swans in population dynamics highlights the importance of developing robust conservation and management strategies --- particularly as the frequency and magnitude of climatic extremes increase over the next century.

@@

Black swans are unexpected extreme events with potentially dramatic consequences \citep{taleb2007,sornette2009}. One of the most striking black swans in ecology is the possible asteroid collision marking the end-Cretaceous mass extinction 65 million years ago \citep{alvarez1980,harnik2012}. Today, climate extremes, in concert with shifts in mean temperature, are expected to cause the greatest ecological and societal damage \citep{ipcc2012}. But, while extremes in the physical environment such as wave height, storm severity, and temperature are frequent experiences \citep{gaines1993,katz2005}, it remains unclear the extent to which ecological systems buffer or suffer from black swans \citep{nunez2012}.

@@

There are two key reasons why we may find little evidence of ecological black swans. First, they might not exist in higher taxa. Indeed, the majority of model fitting and risk forecasting assumes that population dynamics are not heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively, heavy-tailed dynamics might exist, but our ability to detect them requires further development of statistical tools. One such tool is the generalized extreme value distribution, which has been applied to environmental data \citep[e.g.][]{katz2005}. This distribution describes the most extreme event per time interval (e.g.~heaviest rainfall per year), but requires time series that are sufficiently long to be condensed into time intervals. Another statistical tool involves fitting a catastrophic mixture distribution in a state space model to quantify the probability that population events are extreme, although this is also data intensive \citep{ward2007}. A third tool is to compare the support for fits of thin- and heavy-tailed distributions \citep{halley2002}, but this analysis did not quantify the probability of black swans or allow for population dynamics.

Here we develop a framework for identifying ecological black swans in the process noise of population dynamics, i.e.\ the stochastic jumps in abundance from one time step to the next. Our framework quantifies both the magnitude and probability of heavy-tailed dynamics, allows for a range of population dynamic models, can incorporate observation uncertainty, and can be easily applied to abundance time series. We apply our model to \NPops\ populations and estimate the frequency and strength of black-swan events in population dynamics across taxonomy; verify some causes of black-swan events; and attempt to identify characteristics of time series or intrinsic life-history characteristics that are associated with heavy-tailed dynamics.
%We find a high probability of ecological
%black swans across a wide range of taxa but in a limited number of populations.
%When they occur, extreme climate, predation, parasites, and their interactions
%are common documented causes. We find little evidence that intrinsic
%life-history characteristics are related to the probability of black swans.

@@

To obtain estimates of the probability and magnitude of black-swan events, we fit population dynamic models to abundance time series from around the world. For each population, we estimated the shape of the process noise tails by measuring the degrees of freedom ($\nu$) of a Student t-distribution.

@@

We selected abundance time series from the Global Population Dynamics Database (GPDD; \citeauthor{gpdd2010} \citeyear{gpdd2010}), which contains nearly 5000 time series of abundance from $\sim$1000 species and $\sim$100 taxonomic orders. We filtered the data (Supporting Information) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values \citep[\textit{sensu}][]{brook2006a}. Our interpolation affected only $\sim$\interpPointsPerc \% of the final data points (Table~\ref{tab:stats}) and none of the data points that were later considered black-swan events. Our final dataset contained \NPops~populations across \NOrders~taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps~time steps (range of \minTimeSteps--\maxTimeSteps) (Table~\ref{tab:stats}, Fig.~\ref{fig:all-ts}).

@@

By allowing the process noise to be drawn from a Student-t distribution we can estimate the degree to which the process errors have heavy tails and are thereof evidence of black-swan events (Fig.~\ref{fig:didactic}a, b). For example, at $\nu = 3$, the probability of drawing a value more than five standard deviations below the mean is $0.02$, whereas the probability of drawing such a value from a normal distribution is tiny ($2.9\cdot10^{-7}$). As the value of $\nu$ approaches infinity, the distribution approaches the normal distribution (Fig.~\ref{fig:didactic}a, b).

@@

We chose weakly informative priors to incorporate our understanding of plausible population dynamics (\citeauthor{gelman2014} \citeyear{gelman2014}; Supporting Information). For $\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior than suggested by \citet{fernandez1998}. This prior gives only a \basePriorProbHeavy \% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity. In any case, for $\nu > 20$ the t distribution is almost indistinguishable from the normal distribution (Fig.~\ref{fig:didactic}). We investigated the sensitivity of our results to weaker and stronger priors (Fig.~\ref{fig:priors}, Supporting Information). We used simulated data to test how easily we could detect $\nu$ given different sample sizes and to ensure we could recover unbiased parameter estimates from the Gompertz model (Supporting Information).

@@

% "Perhaps then you could say "we examined all time series with published explanations of why the black swan events occurred"
TODO STATE HOW THESE 7 WERE CHOSEN
TODO QUANTIFY UPS AND DOWNS --- SKEW
We assembled the documented causes of black-swan population dynamics in seven exemplar populations with high probabilities of heavy tails (Table~\ref{tab:sparks}). The majority of documented events were downward black swans and involved a series of unfortunate events. For example, a shortage of nest sites for the shag long-necked bird (TODO SCI.\ NAME) in the United Kingdom reduced population productivity and made the population more vulnerable when a red tide hit in 1968 \citep{potts1980}. The population experienced a black-swan downswing, but the new availability of nest sites resulted in a rapid population upswing \citep{potts1980}. As another example, an interaction between environmental- and predation-mediated population cycles are thought to have caused a downward black swan for a water vole population \citep{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For example, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron abundance in the United Kingdom \citep{stafford1971}. Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted \citep{stafford1971}.

@@

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations --- the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find qualitatively similar evidence for black swans in many other large taxonomic or geographic samples of populations, the common forces driving those black swans likely differ (e.g.~severe winters in Table~\ref{tab:sparks}). In addition to a possibly biased sample of populations, some black-swan detections could just be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. A final caveat is that the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black swans. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

@@

The following supporting information is available online for this article:\\
Tables S1 and S2\\
Figures S1--S6\\
Example Stan code for a heavy-tailed Gompertz model and multilevel beta
regression\\
The GPDD IDs used in our analysis\\
All code and data to recreate our analysis are available at:\\
\url{https://github.com/seananderson/heavy-tails}

### 2014-11-08 text/supp.tex
Edits to text

\noindent
We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated also only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}). Note that interpolation is highly unlikely to lead to black-swan detections, since black swans involve extreme increases or decreases. Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.

@@

\clearpage
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\input{../analysis/stat-table}
\clearpage

>>>>

\noindent
We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated also only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}). Note that interpolation is highly unlikely to lead to black-swan detections, since black swans involve extreme increases or decreases. Table~\ref{tab:stats} shows the final taxonomic breakdown and the number of populations with interpolated values.

@@

\clearpage
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\begin{table}
\begin{footnotesize}
\caption{Summary statistics for the filtered Global Population Dynamics Database time series arranged by taxonomic class. Columns are: number of populations, number of taxonomic orders, numbers of species, median time series length, total number of interpolated time steps, total number of substituted zeros, and total number of time steps.}
\smallskip
\begin{tabular}{lrrrrrrrr}
\toprule
\input{../analysis/stat-table.tex}
\label{tab:stats}
\end{tabular}
\end{footnotesize}
\end{table}
